### 2017-01-16 2017mergingmerges/section/XX-Fusion.tex
rejigging syntax etc

Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Fusion proceeds by taking two processes in a nest, and creating a single process that computes both.
The new process uses labels made of the product of the two input processes, as well as the static part of the input state - that is, whether the input buffer is pending after injection, has been pulled, or is empty, but without the actual injected value.

@@

Before looking at the definitions, it is worth noting the types and purposes of important functions:
\begin{itemize}
\item
\ti{fuseNest} takes a process nest and repeatedly fuses pairs together. 
It chooses arbitrary processes with shared channels and fuses them together.

>>>>

Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
Fusion proceeds by taking two processes in a network, and creating a single process that computes both.
The new process uses labels made of the product of the two input processes, as well as the static part of the input state - that is, whether the input buffer is pending after injection, has been pulled, or is empty, but without the actual injected value.

@@

Before looking at the definitions, it is worth noting the types and purposes of important functions:
\begin{itemize}
\item
\ti{fuseNetwork} takes a process network and repeatedly fuses pairs together. 
It chooses arbitrary processes with shared channels and fuses them together.

### 2017-01-16 2017mergingmerges/section/XX-FutureWork.tex
rejigging syntax etc

We add two new instructions for closing output streams and disconnecting from input streams.
Closing an output stream $(@close@~\Chan~\Goto)$ is similar to pushing an end-of-file marker to all readers.
As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
After a stream has been closed, no further values can be pushed.

Disconnecting from input streams $(@disconnect@~\Chan~\Goto)$ signals that a process is no longer interested in the values of a stream.
This can be used when a process requires the first values of a stream, but does not require the whole stream.
If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
After a process disconnects from an input channel, it can no longer pull from that channel.

>>>>

We add two new instructions for closing output streams and disconnecting from input streams.
Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
After a stream has been closed, no further values can be pushed.

Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
This can be used when a process requires the first values of a stream, but does not require the whole stream.
If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
After a process disconnects from an input channel, it can no longer pull from that channel.

### 2017-01-16 2017mergingmerges/section/XX-Processes.tex
prose
rejigging syntax etc

All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
Combining variable update with stream instructions simplifies the fusion process~\REF, as some parts of fusion need to perform both at once.

A process nest is a set of multiple processes that can be evaluated concurrently.
The intersection of all process outputs should be empty - there should be no overlap.
Any inputs that are not mentioned as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

One of the simplest combinators is @map@.
This might need to go elsewhere.
As well as needing a better example than map/map.
Let's start with the process definition for @map@.
The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
The goto labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
Mention that the initial heap has the names but no values, but they could be initialised to whatever.
It doesn't matter since they'll be written before anything is read.

@@

As well as a ``combinator nest'', a function comprised exclusively of process combinators.
The input streams are supplied as arguments, and output streams as return values.
\begin{code}
mapMap f g xs
 = let ys = map f xs
       zs = map g ys
   in  zs
\end{code}

It is not hard to assume that, given the process definitions and a combinator nest, we can produce a process nest.
This is simple enough for a paragraph prose description.
It's just a bit of inlining and renaming everything to be unique.

@@

Now we can perform some kind of fusion on this nest, resulting in one process that computes, as output, both @ys@ and @zs@.
Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator nest did not return them.

@@

Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall nest's input streams.
For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.

@@

(ProcessesInject) perform injection over a process nest.
Every process in the nest must have the value injected into it.
This means if multiple processes read from that stream, all input buffers for that stream must be empty.

@@

The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{\phi'}$
executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
The output message $\alpha$ can be an internal state change or an emitted value.
The result also has the new label, the new input state buffers, and the substitution to apply to the heap.

The two judgment forms for shaking processes are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p}}$.
The process shaking just shakes a single instruction and updates the process.
Shaking a process nest chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the nest.

(Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
The result substitution sets the variable to the pulled value, as well as any substitutions in the \Goto~ of the instruction.

@@

(Shake) unwraps a single process and evaluates the instruction.
It updates the process with the new label, input state and heap.

(ProcessesInternal) chooses one process from the nest and evaluates it.
When the process evaluates with an internal message ($\tau$), the entire nest evaluates by replacing that process.

(ProcessesPush) chooses one process from the nest and evaluates it, where the process evaluates with a push message.
The emitted push message is then injected into all other processes in the nest, which means they must either ignore the channel or be ready to add it to their buffer.
If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
The entire nest then emits the same message.

The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process nest and output values are accumulated.
The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{nest}}{\ti{streams}}{\ti{nest}'}$.
The input map $\ti{inputs}$ contains values for the nest inputs: nest outputs are not allowed, but ignored channels can have values.
The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
Feeding evaluates the process nest until all input values have been injected.

Note that the result stream and nest are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.

@@

(FeedInternal) first recursively feeds its input accumulator and process nest, then allows the resulting nest to take an internal step.
The internal step does not affect the accumulators.
The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.

(FeedPush) works similarly to (FeedInternal) except that the process nest emits a push message.
The pushed value is added to the end of the accumulator for that channel.

(FeedExternal) allows inputs to be injected into the process nest.
For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
The recursive feed is evaluated with the last value removed from the accumulators.
The last value is then injected into the nest, and added back to the result accumulators.

>>>>

All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.

A process network is a set of multiple processes that can be evaluated concurrently.
The intersection of all process outputs should be empty - there should be no overlap.
Any inputs that are not mentioned as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

@@

One of the simplest combinators is @map@.
This might need to go elsewhere.
As well as needing a better example than map/map.
Let's start with the process definition for @map@.
The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
Mention that the initial heap has the names but no values, but they could be initialised to whatever.
It doesn't matter since they'll be written before anything is read.

@@

As well as a ``combinator network'', a function comprised exclusively of process combinators.
The input streams are supplied as arguments, and output streams as return values.
\begin{code}
mapMap f g xs
 = let ys = map f xs
       zs = map g ys
   in  zs
\end{code}

It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
This is simple enough for a paragraph prose description.
It's just a bit of inlining and renaming everything to be unique.

@@

Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.

@@

Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall network's input streams.
For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.

@@

(ProcessesInject) performs injection over a process network.
Every process in the network must have the value injected into it.
This means if multiple processes read from that stream, all input buffers for that stream must be empty.

@@

The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
The output message $\alpha$ can be an internal state change or an emitted value.
The result also has the new label, the new input state buffers, and the substitution to apply to the heap.

The two judgment forms for shaking processes are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p}}$.
The process shaking just shakes a single instruction and updates the process.
Shaking a process network chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the network.

(Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.

@@

(Shake) unwraps a single process and evaluates the instruction.
The instruction updates are evaluated and updated in the process heap.
It updates the process with the new label, input state and heap.

(ProcessesInternal) chooses one process from the network and evaluates it.
When the process evaluates with an internal message ($\tau$), the entire network evaluates by replacing that process.

(ProcessesPush) chooses one process from the network and evaluates it, where the process evaluates with a push message.
The emitted push message is then injected into all other processes in the network, which means they must either ignore the channel or be ready to add it to their buffer.
If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
The entire network then emits the same message.

The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process network and output values are accumulated.
The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
Feeding evaluates the process network until all input values have been injected.

Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.

@@

(FeedInternal) first recursively feeds its input accumulator and process network, then allows the resulting network to take an internal step.
The internal step does not affect the accumulators.
The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.

(FeedPush) works similarly to (FeedInternal) except that the process network emits a push message.
The pushed value is added to the end of the accumulator for that channel.

(FeedExternal) allows inputs to be injected into the process network.
For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
The recursive feed is evaluated with the last value removed from the accumulators.
The last value is then injected into the network, and added back to the result accumulators.

### 2017-01-16 2017mergingmerges/section/XX-Proof.tex
rejigging syntax etc

Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair} if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
It is important to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
This is because evaluation of a process nest is non-deterministic, and fusion commits to a particular evaluation order.

The problem with commiting to an evaluation order is most easily explained with a process nest containing two infinitely pushing processes.
Process @A@ is repeatedly pushing to a stream called @X@, while process @B@ repeatedly pushes to @Y@.
When evaluating this pair as a process nest, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.

@@

The system described here has some differences to our Coq formalisation.
First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Goto~label of any instruction.
This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
Ideally, a future version of the formalisation would have this change.
Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
Instead we sequentially evaluate both processes separately with the same input values and outputs.
Finally, we have not implemented \ti{fuseNest} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.

>>>>

Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair}: if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
It is interesting to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
This is because evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.

The problem with commiting to an evaluation order is most easily explained with a process network containing two infinitely pushing processes.
Process @A@ is repeatedly pushing to a stream called @X@, while process @B@ repeatedly pushes to @Y@.
When evaluating this pair as a process network, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.

@@

The system described here has some differences to our Coq formalisation.
First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Next~label of any instruction.
This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
Ideally, a future version of the formalisation would have this change.
Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
Instead we sequentially evaluate both processes separately with the same input values and outputs.
Finally, we have not implemented \ti{fuseNetwork} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.
